GMSE: an R package for generalised management strategy evaluation

### 2017-11-07 notebook/ms/SI1.Rmd
Introduction to the process

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decions that all other users follow identically. Each run of the genetic algorithm mimics the evolution by natural selection of a populaiton of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text). Below, we explain the genetic algorithm in extended detail as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework). We first explain the key data structure used, then provide an overview of the processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement. We then explain the fitness functions of managers and users in more detail.

@@

The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`). Rows of `ACTION` correspond to recipients of actions (resources, landscapes, or potentially other agents), and columns correspond to either recipient properties or actions allocated to recipients. Each layer of `ACTION` corresponds to a unique agent, the first of which is the manager; additional layers correspond to each user. Below shows an `ACTION` array, which corresponds to a GMSE model with one manager and two users.

@@

The above array holds all of the information on manager and user actions in the six right-most columns of each array layer. The first seven columns contain information about which resources are affected, and how they are affected. The first column `Act` identifies the type of action being performed; a value of -2 defines a direct action to a resource (e.g., culling of the resource), and a value of -1 defines direct action to a landscape (e.g., increasing yield). Positive values are currently only meaningful for `Manager_Actions`, where a value of 1 defines an action setting a uniform cost of users' direct actions on resources (i.e., costs where `Act = -2` for `User_1_Actions` and `User_2_Actions`). All other values for `Act` are meaningless in GMSE 0.3.1.7, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation). Similarly, columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in GMSE v0.3.1.7 (i.e., only one type of resource is permitted), but future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). For the rest of this supporting information, we will therefore focus only on rows 1-3 of `ACTION`. Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identifies whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the marginal utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Res_cost = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.

Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`) of dimensions identical to `ACTION`. All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible budget of either managers or users, so neither can affect resource types or utilities. Columns 8-13 are also set to 10001, except where actions are specifically allowed. Below shows the `COST` array that corresponds to the above `ACTION` array.

@@

Note that in default GMSE parameters, `culling = TRUE`, but all other actions are false. Hence the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 10001. Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12). This can be verified in `ACTION` where the manager has set the cost of scaring to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users. The cost of the manager affecting the cost of user actions is always set to the `minimum_cost` defined in GMSE; here the default `r samp_COST[3,9,1]` is used. This `minimum_cost` also defines cost values for `None`, in which the user or manager does nothing, as might occur if the manager wants to permit culling and therefore does not want to invest any of their `manager_budget` to increasing the cost of culling. Both `ACTION` and `COST` are updated in each time step unless `manage_freq > 1`, in which case `COST` and `Manager_Actions` in `ACTION` are at the frequency defined.

@@

It is the `ACTION` array that is affected by the genetic algorithm.

>>>>

A genetic algorithm is called in the predefined GMSE manager and user models to simulate human decision making. As of GMSE version 0.3.1.9, this includes one independent call to the genetic algorithm for each decision-making agent in every GMSE time step. Therefore, one run of the genetic algorithm occurs to simulate the manager's policy setting decisions in each time step (unless otherwise defined through non-default `manage_freq` values greater than 1), and one run occurs to simulate each individual user's action decisions in each time step (unless otherwise defined through non-default `group_think = TRUE`, in which case one user makes decisions that all other users follow identically. Each run of the genetic algorithm mimics the evolution by natural selection of a population of potential manager or user strategies over multiple generations, with the highest fitness strategy in the terminal generation being selected as the one that the manager or user decides to implement. For clarity, as in the main text, we use 'time step' to refer to a full GMSE cycle (in which multiple genetic algorithms may be run) and 'generation' to refer to a single, non-overlapping, generation of potential strategies that evolve within a genetic algorithm (see Figure 1 of the main text). Below, we explain the genetic algorithm in detail, as it occurs in GMSE v0.3.1.9 (future versions of GMSE might expand upon this framework). We first explain the key data structures used, then provide an overview of how a population of strategies is initialised, and the  subsequent processes of crossover, mutation, cost constraint, fitness evaluation, tournament selection, and replacement. We then explain the fitness functions of managers and users in more detail.

@@

The focal data structure used for tracking manager and user decisions is a three dimensional array, which we will call `ACTION` (also returned as `user_array` by `gmse_apply`). Rows of `ACTION` correspond to the entities affected by actions (resources, landscape properties, or potentially other agents), and columns correspond either to properties of the affected entities, or to the actions potentially allocated to them. Each layer of `ACTION` corresponds to a unique agent, the first of which is the manager; additional layers correspond to users. Below shows an `ACTION` array for a GMSE model with one manager and two users.

@@

The above array holds all of the information on manager and user actions. The first seven columns contain information about which entities are affected, and how they are affected. The first column `Act` identifies the type of action being performed; a value of -2 defines a direct action to a resource (e.g., culling of the resource), and a value of -1 defines direct action to a landscape (e.g., increasing yield). Positive values are currently only meaningful for `Manager_Actions`, where a value of 1 defines an action setting a uniform cost of users' direct actions on resources (i.e., costs where `Act = -2` for `User_1_Actions` and `User_2_Actions`). All other values for `Act` are meaningless in GMSE 0.3.1.9, but might be expanded upon in future versions to allow for modification of specific user costs enacted by managers (i.e., managers having different policies for different users) or other users (e.g., users increasing the costs of other users' actions due to conflict or cooperation). Similarly, columns 2-4 refer to resource or landscape types, but only `Type_1 = 1`, `Type_2 = 0`, and `Type_3 = 0 ` are allowed in GMSE v0.3.1.9 (i.e., only one type of resource is permitted). Future versions might allow for different resource types (e.g., `Type_1` might be used to designate species, and `Type_2` and `Type_3` could designate stage or sex). For the rest of this supporting information, we will therefore focus only on rows 1-3 of `ACTION`. Column 5 `Util.` of `ACTION` defines the utility associated with the resource (where `Act = -2`) or landscape (where `Act = -1`). For managers, the target resource abundance set with the GMSE argument `manage_target` is found in row 1 (`r samp_ACTION[1,5,1]` in `ACTION` above); for users, the value in row 1 identifies whether resources are preferred to increase (if positive) or decrease (if negative). Values of column 5 in row 2 similarly identify whether landscape cell output is preferred by users to increase or decrease (managers do not currently have preferences for landscape output). Of special note is row 3 for `Manager_Actions`, which defines the marginal utility of resources; that is, the adjustment to resource abundance that the manager will attempt to make based on the `manage_target` and the estimated abundance produced by the observation model (in the case of the above, resource abundance is estimated at ca `r round((samp_ACTION[1,5,1] + -1*samp_ACTION[3,5,1]),2)`, so the manager will set policy in attempt to change the population size by ca `r round(samp_ACTION[3,5,1],2)` resources). Column 6 `U_land` defines whether or not the utility attached to the resource or landscape output depends on it being on a landscape cell that is owned by the acting user. Related, column 7 `U_loc.` defines whether or not actions can be performed only on a landscape cell that is owned by the acting user. Hence values of columns 6 and 7 are binary, and affected by the `land_ownership` argument in `gmse`. Finally, columns 8-13 correspond to specific actions, either direct (where `Act < 0`) or indirect by setting policy (for row 3 of `Manager_Actions` where `Act = 1`). The last column 13 `None` corresponds with no actions. See [GMSE documentation](https://cran.r-project.org/web/packages/GMSE/GMSE.pdf) for details about the effects of each action.

Constraints on the values that elements in the `ACTION` array can take are defined by a `COST` array (also returned as `manager_array` by `gmse_apply`) of dimensions identical to `ACTION`. Elements of `COST` define how many units from the `manager_budget` or `user_budget` are needed to perform a single action; a `minimum_cost` for actions is defined as an argument in GMSE (10 by default). All values in `COST` columns 1-7 are set to 10001, one higher than the highest possible `manager_budget` or `user_budget`, so neither can affect resource types or utilities. Columns 8-13 are also set to 10001, except where actions are allowed. Below shows the `COST` array that corresponds to the above `ACTION` array.

@@

Note that in default GMSE parameters, `culling = TRUE`, but all other actions are false. Hence the `Cull` column 9 is the only column besides column 13 `None` in which cost is less than 10001. Manager's actions in `ACTION` directly affect the cost of users performing one of the five possible actions on resources (columns 8-12). This can be verified in `ACTION` where the manager has set the cost of scaring to `r samp_ACTION[3,9,1]`, and the corresponding `COST` of resource culling (row 1) is `r samp_COST[1,9,2]` for both users. The cost of the manager affecting the cost of user actions is always set to the `minimum_cost`; here the default `r samp_COST[3,9,1]` is used. This `minimum_cost` also defines cost values for `None`, in which the user or manager does nothing, as might occur if the manager wants to permit culling and therefore does not want to invest any of their `manager_budget` to increasing the cost of culling. Both `ACTION` and `COST` are updated in each time step unless `manage_freq > 1`, in which case `COST` and `Manager_Actions` in `ACTION` are updated at the frequency defined.

@@

The genetic algorithm updates a single layer of the `ACTION` array, which defines to the decisions of a single agent (either the manager or a user). The corresponding layer of the `COST` array remains unchanged, and serves only to ensure that `ACTION` values do not exceed `manager_budget` or `user_budget` for managers and users, respectively. The genetic algorithm proceeds by first initialising a large population of new `ACTION` layers. In each generation, these layers crossover and mutate, generating variation in agent decisions; costs constrain this variation from exceeding a maximum budget, then the fitness of each layer is evaluated based on how much it increases the agent's utility. A tournament is used to select high fitness layers, and these selected layers become the new generation of layers; generations continue until a minimum number of generations (`ga_mingen`) have passed and a convergence criteria is satisfied such that the change in mean fitness from the previous generation is below the threshold `converge_crit`.

Initialisation
--------------------------------------------------------------------------------
